---
title: "Lesson 13"
author: "Christopher A. Swenson (chris@cswenson.com)"
date: "11/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Galton peas (nonconstant variance and weighted least squares)
Load the galton data.
Fit an ordinary least squares (OLS) simple linear regression model of Progeny vs Parent.
Fit a weighted least squares (WLS) model using weights = .
Create a scatterplot of the data with a regression line for each model.
```{r galton}
galton <- read.table("./Data/galton.txt", header=T)
attach(galton)
 
model.1 <- lm(Progeny ~ Parent)
summary(model.1)
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept) 0.127029   0.006993  18.164 9.29e-06 ***
# Parent      0.210000   0.038614   5.438  0.00285 ** 
 
model.2 <- lm(Progeny ~ Parent, weights=1/SD^2)
summary(model.2)
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept) 0.127964   0.006811  18.787 7.87e-06 ***
# Parent      0.204801   0.038155   5.368  0.00302 ** 
 
plot(x=Parent, y=Progeny, ylim=c(0.158,0.174),
     panel.last = c(lines(sort(Parent), fitted(model.1)[order(Parent)], col="blue"),
                    lines(sort(Parent), fitted(model.2)[order(Parent)], col="red")))
legend("topleft", col=c("blue","red"), lty=1,
       inset=0.02, legend=c("OLS", "WLS"))
 
detach(galton)
```

## Computer-assisted learning (nonconstant variance and weighted least squares)
Load the ca_learning data.
Create a scatterplot of the data.
Fit an OLS model.
Plot the OLS residuals vs num.responses.
Plot the absolute OLS residuals vs num.responses.
Calculate fitted values from a regression of absolute residuals vs num.responses.
Fit a WLS model using weights = .
Create a scatterplot of the data with a regression line for each model.
Plot the WLS standardized residuals vs num.responses.
```{r computer}
ca_learning <- read.table("./Data/ca_learning.txt", header=T)
attach(ca_learning)
 
plot(x=num.responses, y=cost)
 
model.1 <- lm(cost ~ num.responses)
summary(model.1)
#               Estimate Std. Error t value Pr(>|t|)    
# (Intercept)    19.4727     5.5162   3.530  0.00545 ** 
# num.responses   3.2689     0.3651   8.955 4.33e-06 ***
# ---
# Residual standard error: 4.598 on 10 degrees of freedom
# Multiple R-squared:  0.8891,  Adjusted R-squared:  0.878 
# F-statistic: 80.19 on 1 and 10 DF,  p-value: 4.33e-06
 
plot(num.responses, residuals(model.1))
plot(num.responses, abs(residuals(model.1)))
 
wts <- 1/fitted(lm(abs(residuals(model.1)) ~ num.responses))^2
 
model.2 <- lm(cost ~ num.responses, weights=wts)
summary(model.2)
#               Estimate Std. Error t value Pr(>|t|)    
# (Intercept)    17.3006     4.8277   3.584  0.00498 ** 
# num.responses   3.4211     0.3703   9.238 3.27e-06 ***
# ---
# Residual standard error: 1.159 on 10 degrees of freedom
# Multiple R-squared:  0.8951,  Adjusted R-squared:  0.8846 
# F-statistic: 85.35 on 1 and 10 DF,  p-value: 3.269e-06
 
plot(x=num.responses, y=cost, ylim=c(50,95),
     panel.last = c(lines(sort(num.responses), fitted(model.1)[order(num.responses)], col="blue"),
                    lines(sort(num.responses), fitted(model.2)[order(num.responses)], col="red")))
legend("topleft", col=c("blue","red"), lty=1,
       inset=0.02, legend=c("OLS", "WLS"))
 
plot(num.responses, rstandard(model.2))
 
detach(ca_learning)
```

## Market share (nonconstant variance and weighted least squares)
Load the marketshare data.
Fit an OLS model.
Plot the OLS residuals vs fitted values with points marked by Discount.
Use the tapply function to calculate the residual variance for Discount=0 and Discount=1.
Fit a WLS model using weights = 1/variance for Discount=0 and Discount=1.
Plot the WLS standardized residuals vs fitted values.
```{r ms}
marketshare <- read.table("./Data/market_share.txt", header=T)
attach(marketshare)
 
model.1 <- lm(MarketShare ~ Price + P1 + P2)
summary(model.1)
  #             Estimate Std. Error t value Pr(>|t|)    
# (Intercept)  3.19592    0.35616   8.973 3.00e-10 ***
# Price       -0.33358    0.15226  -2.191   0.0359 *  
# P1           0.30808    0.06412   4.804 3.51e-05 ***
# P2           0.48431    0.05541   8.740 5.49e-10 ***
 
plot(fitted(model.1), residuals(model.1), col=Discount+1)
vars <- tapply(residuals(model.1), Discount, var)
#          0          1 
# 0.01052324 0.02680546 
 
wts <- Discount/vars[2] + (1-Discount)/vars[1]
 
model.2 <- lm(MarketShare ~ Price + P1 + P2, weights=wts)
summary(model.2)
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept)  3.17437    0.35671   8.899 3.63e-10 ***
# Price       -0.32432    0.15291  -2.121   0.0418 *  
# P1           0.30834    0.06575   4.689 4.89e-05 ***
# P2           0.48419    0.05422   8.930 3.35e-10 ***
 
plot(fitted(model.2), rstandard(model.2), col=Discount+1)
 
detach(marketshare)
```

## Home price (nonconstant variance and weighted least squares)
Load the realestate data.
Calculate log transformations of the variables.
Fit an OLS model.
Plot the OLS residuals vs fitted values.
Calculate fitted values from a regression of absolute residuals vs fitted values.
Fit a WLS model using weights = .
Plot the WLS standardized residuals vs fitted values.
```{r real}
realestate <- read.table("./Data/realestate.txt", header=T)
attach(realestate)
 
logY <- log(SalePrice)
logX1 <- log(SqFeet)
logX2 <- log(Lot)
 
model.1 <- lm(logY ~ logX1 + logX2)
summary(model.1)
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept)  4.25485    0.07353  57.864  < 2e-16 ***
# logX1        1.22141    0.03371  36.234  < 2e-16 ***
# logX2        0.10595    0.02394   4.425 1.18e-05 ***
 
plot(fitted(model.1), residuals(model.1))
 
wts <- 1/fitted(lm(abs(residuals(model.1)) ~ fitted(model.1)))^2
 
model.2 <- lm(logY ~ logX1 + logX2, weights=wts)
summary(model.2)
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept)  4.35189    0.06330  68.755  < 2e-16 ***
# logX1        1.20150    0.03332  36.065  < 2e-16 ***
# logX2        0.07924    0.02152   3.682 0.000255 ***
 
plot(fitted(model.2), rstandard(model.2))
 
detach(realestate)
```
