---
title: "Lesson 08"
author: "Christopher A. Swenson (chris@cswenson.com)"
date: "11/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Birthweight and smoking (2-level categorical predictor, additive model)
Load the birthsmokers data.
Create a scatterplot matrix of the data
Fit a multiple linear regression model of Wgt on Gest + Smoke.
Display scatterplot of Wgt vs Gest with points marked by Smoke and add parallel regression lines representing Smoke=0 and Smoke=1.
Display regression results and calculate confidence intervals for the regression parameters.
Display confidence intervals for expected Wgt at Gest=38 (for Smoke=1 and Smoke=0).
Repeat analysis separately for Smoke=0 and Smoke=1.
Repeat analysis using (1, -1) coding.
```{r smoke}
birthsmokers <- read.table("./Data/birthsmokers.txt", header=T)
attach(birthsmokers)
 
pairs(cbind(Wgt, Gest, Smoke))
 
model <- lm(Wgt ~ Gest + Smoke)
 
plot(x=Gest, y=Wgt, ylim=c(2300, 3700), 
     col=ifelse(Smoke=="yes", "red", "blue"),
     panel.last = c(lines(sort(Gest[Smoke=="no"]),
                          fitted(model)[Smoke=="no"][order(Gest[Smoke=="no"])],
                          col="blue"),
                    lines(sort(Gest[Smoke=="yes"]),
                          fitted(model)[Smoke=="yes"][order(Gest[Smoke=="yes"])],
                          col="red")))
 
summary(model)
#              Estimate Std. Error t value Pr(>|t|)    
# (Intercept) -2389.573    349.206  -6.843 1.63e-07 ***
# Gest          143.100      9.128  15.677 1.07e-15 ***
# Smokeyes     -244.544     41.982  -5.825 2.58e-06 ***
# ---
# Residual standard error: 115.5 on 29 degrees of freedom
# Multiple R-squared:  0.8964,  Adjusted R-squared:  0.8892 
# F-statistic: 125.4 on 2 and 29 DF,  p-value: 5.289e-15
 
confint(model)
#                  2.5 %     97.5 %
# (Intercept) -3103.7795 -1675.3663
# Gest          124.4312   161.7694
# Smokeyes     -330.4064  -158.6817
 
predict(model, interval="confidence",
        newdata=data.frame(Gest=c(38, 38), Smoke=c(1, 0)))
#        fit      lwr      upr
# 1 2803.693 2740.599 2866.788
# 2 3048.237 2989.120 3107.355
 
model.0 <- lm(Wgt ~ Gest, subset=Smoke==0)
summary(model.0)
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept) -2546.14     457.29  -5.568 6.93e-05 ***
# Gest          147.21      11.97  12.294 6.85e-09 ***
 
predict(model.0, interval="confidence",
        newdata=data.frame(Gest=38))
#        fit      lwr     upr
# 1 3047.724 2990.298 3105.15
 
model.1 <- lm(Wgt ~ Gest, subset=Smoke==1)
summary(model.1)
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept) -2474.56     553.97  -4.467 0.000532 ***
# Gest          139.03      14.11   9.851 1.12e-07 ***
 
predict(model.1, interval="confidence",
        newdata=data.frame(Gest=38))
#        fit      lwr      upr
# 1 2808.528 2731.726 2885.331
 
Smoke2 <- ifelse(Smoke==1, 1, -1)
model.3 <- lm(Wgt ~ Gest + Smoke2)
summary(model.3)
#              Estimate Std. Error t value Pr(>|t|)    
# (Intercept) -2511.845    353.449  -7.107 8.07e-08 ***
# Gest          143.100      9.128  15.677 1.07e-15 ***
# Smoke2       -122.272     20.991  -5.825 2.58e-06 ***
 
# Alternatively
#model.3 <- lm(Wgt ~ Gest + Smoke, contrasts=list(Smoke="contr.sum"))
model.3 <- lm(Wgt ~ Gest + Smoke)
summary(model.3)
 
detach(birthsmokers)
```

## Depression treatments (3-level categorical predictor, interaction model)
Load the depression data.
Display scatterplot of y (treatment effectiveness) vs age with points marked by treatment.
Create interaction variables and fit a multiple linear regression model of y on age + x2 + x3 + age.x2 + age.x3.
Add non-parallel regression lines representing each of the three treatments to the scatterplot.
Display a residuals vs fits plot and a normal probability plot of the residuals, and conduct an Anderson-Darling normality test using the nortest package.
Conduct an F-test to see if at least one of x2, x3, age.x2, and age.x3 are useful (i.e., the regression functions differ).
Conduct an F-test to see if at least one of age.x2 and age.x3 are useful (i.e., the regression functions have different slopes).
```{r dep}
depression <- read.table("./Data/depression.txt", header=T)
attach(depression)
 
plot(x=age, y=y, col=as.numeric(TRT))
legend("topleft", col=1:3, pch=1,
       inset=0.02, x.intersp = 1.5, y.intersp = 1.8,
       legend=c("Trt A", "Trt B", "Trt C"))
 
age.x2 <- age*x2
age.x3 <- age*x3
 
model.1 <- lm(y ~ age + x2 + x3 + age.x2 + age.x3)
summary(model.1)
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept)  6.21138    3.34964   1.854 0.073545 .  
# age          1.03339    0.07233  14.288 6.34e-15 ***
# x2          41.30421    5.08453   8.124 4.56e-09 ***
# x3          22.70682    5.09097   4.460 0.000106 ***
# age.x2      -0.70288    0.10896  -6.451 3.98e-07 ***
# age.x3      -0.50971    0.11039  -4.617 6.85e-05 ***
 
plot(x=age, y=y, ylim=c(20, 80), col=as.numeric(TRT),
     panel.last = c(lines(sort(age[TRT=="A"]),
                          fitted(model.1)[TRT=="A"][order(age[TRT=="A"])],
                          col=1),
                    lines(sort(age[TRT=="B"]),
                          fitted(model.1)[TRT=="B"][order(age[TRT=="B"])],
                          col=2),
                    lines(sort(age[TRT=="C"]),
                          fitted(model.1)[TRT=="C"][order(age[TRT=="C"])],
                          col=3)))
legend("topleft", col=1:3, pch=1, lty=1,
       inset=0.02, x.intersp = 1.5, y.intersp = 1.8,
       legend=c("Trt A", "Trt B", "Trt C"))
 
plot(x=fitted(model.1), y=rstudent(model.1),
     panel.last = abline(h=0, lty=2))
 
qqnorm(residuals(model.1), main="", datax=TRUE)
qqline(residuals(model.1), datax=TRUE)
 
library(nortest)
ad.test(residuals(model.1)) # A = 0.4057, p-value = 0.3345
 
anova(model.1)
#           Df Sum Sq Mean Sq  F value    Pr(>F)    
# age        1 3424.4  3424.4 222.2946 2.059e-15 ***
# x2         1  803.8   803.8  52.1784 4.857e-08 ***
# x3         1    1.2     1.2   0.0772    0.7831    
# age.x2     1  375.0   375.0  24.3430 2.808e-05 ***
# age.x3     1  328.4   328.4  21.3194 6.850e-05 ***
# Residuals 30  462.1    15.4                       
 
model.2 <- lm(y ~ age)
anova(model.2, model.1)
# Model 1: y ~ age
# Model 2: y ~ age + x2 + x3 + age.x2 + age.x3
#   Res.Df     RSS Df Sum of Sq     F    Pr(>F)    
# 1     34 1970.57                                 
# 2     30  462.15  4    1508.4 24.48 4.458e-09 ***
 
model.3 <- lm(y ~ age + x2 + x3)
anova(model.3, model.1)
# Model 1: y ~ age + x2 + x3
# Model 2: y ~ age + x2 + x3 + age.x2 + age.x3
#   Res.Df     RSS Df Sum of Sq      F   Pr(>F)    
# 1     32 1165.57                                 
# 2     30  462.15  2    703.43 22.831 9.41e-07 ***
 
detach(depression)
```

## Real estate air conditioning (2-level categorical predictor, interaction model, transformations)
Load the realestate data.
Create an interaction variable and fit a multiple linear regression model of SalePrice on SqFeet + Air + SqFeet.Air.
Display scatterplot of SalePrice vs SqFeet with points marked by Air and add non-parallel regression lines representing Air=0 and Air=1.
Display residual plot with fitted (predicted) values on the horizontal axis.
Create log(SalePrice), log(SqFeet), and log(SqFeet).Air variables and fit a multiple linear regression model of log(SalePrice) on log(SqFeet) + Air + log(SqFeet).Air.
Display scatterplot of log(SalePrice) vs log(SqFeet) with points marked by Air and add non-parallel regression lines representing Air=0 and Air=1.
Display residual plot with fitted (predicted) values on the horizontal axis.
```{r real}
realestate <- read.table("./Data/realestate_sales.txt", header=T)
attach(realestate)
 
SqFeet.Air <- SqFeet*Air
model.1 <- lm(SalePrice ~ SqFeet + Air + SqFeet.Air)
summary(model.1)
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept)   -3.218     30.085  -0.107 0.914871    
# SqFeet       104.902     15.748   6.661 6.96e-11 ***
# Air          -78.868     32.663  -2.415 0.016100 *  
# SqFeet.Air    55.888     16.580   3.371 0.000805 ***
 
plot(x=SqFeet, y=SalePrice, 
     col=ifelse(Air, "red", "blue"),
     panel.last = c(lines(sort(SqFeet[Air==0]),
                          fitted(model.1)[Air==0][order(SqFeet[Air==0])],
                          col="blue"),
                    lines(sort(SqFeet[Air==1]),
                          fitted(model.1)[Air==1][order(SqFeet[Air==1])],
                          col="red")))
legend("topleft", col=c("blue", "red"), pch=1, lty=1, inset=0.02,
       legend=c("No air conditioning", "Air conditioning"))
 
plot(x=fitted(model.1), y=residuals(model.1),
     xlab="Fitted values", ylab="Residuals",
     panel.last = abline(h=0, lty=2))
 
lnSalePrice <- log(SalePrice)
lnSqFeet <- log(SqFeet)
lnSqFeet.Air <- lnSqFeet*Air
 
model.2 <- lm(lnSalePrice ~ lnSqFeet + Air + lnSqFeet.Air)
 
plot(x=lnSqFeet, y=lnSalePrice, 
     col=ifelse(Air, "red", "blue"),
     panel.last = c(lines(sort(lnSqFeet[Air==0]),
                          fitted(model.2)[Air==0][order(lnSqFeet[Air==0])],
                          col="blue"),
                    lines(sort(lnSqFeet[Air==1]),
                          fitted(model.2)[Air==1][order(lnSqFeet[Air==1])],
                          col="red")))
legend("topleft", col=c("blue", "red"), pch=1, lty=1, inset=0.02,
       legend=c("No air conditioning", "Air conditioning"))
 
plot(x=fitted(model.2), y=residuals(model.2),
     xlab="Fitted values", ylab="Residuals",
     panel.last = abline(h=0, lty=2))
 
detach(realestate)
```

## Hospital infection risk (4-level categorical predictor, additive model)
Load the infectionrisk data and select observations with Stay <= 14.
Create indicator variables for regions.
Fit a multiple linear regression model of InfctRsk on Stay + Xray + i2 + i3 + i4.
Conduct an F-test to see if at least one of i2, i3, and i4 are useful (conclusion: the regression functions differ by region).
Conduct an F-test to see if at least one of i2 and i3 are useful (conclusion: only the west region differs).
```{r infect}
infectionrisk <- read.table("./Data/infectionrisk.txt", header=T)
infectionrisk <- infectionrisk[infectionrisk$Stay<=14,]
attach(infectionrisk)
 
i1 <- ifelse(Region==1,1,0) # NE
i2 <- ifelse(Region==2,1,0) # NC
i3 <- ifelse(Region==3,1,0) # S
i4 <- ifelse(Region==4,1,0) # W
 
model.1 <- lm(InfctRsk ~ Stay + Xray + i2 + i3 + i4)
summary(model.1)
#              Estimate Std. Error t value Pr(>|t|)    
# (Intercept) -2.134259   0.877347  -2.433  0.01668 *  
# Stay         0.505394   0.081455   6.205 1.11e-08 ***
# Xray         0.017587   0.005649   3.113  0.00238 ** 
# i2           0.171284   0.281475   0.609  0.54416    
# i3           0.095461   0.288852   0.330  0.74169    
# i4           1.057835   0.378077   2.798  0.00612 ** 
# ---
# Residual standard error: 1.036 on 105 degrees of freedom
# Multiple R-squared:  0.4198,  Adjusted R-squared:  0.3922 
# F-statistic: 15.19 on 5 and 105 DF,  p-value: 3.243e-11
 
model.2 <- lm(InfctRsk ~ Stay + Xray)
anova(model.2, model.1)
#   Res.Df    RSS Df Sum of Sq      F  Pr(>F)  
# 1    108 123.56                              
# 2    105 112.71  3    10.849 3.3687 0.02135 *
 
model.3 <- lm(InfctRsk ~ Stay + Xray + i4)
anova(model.3, model.1)
#   Res.Df    RSS Df Sum of Sq      F Pr(>F)
# 1    107 113.11                           
# 2    105 112.71  2   0.39949 0.1861 0.8305
 
detach(infectionrisk)
```
