---
title: "Lesson 14"
author: "Christopher A. Swenson (chris@cswenson.com)"
date: "11/27/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Google stock (autoregression model)
Use the read.zoo function in the zoo package to load the google_stock data in time series format.
Create a time series plot of the data.
Load the google_stock data in the usual way using read-table.
Use the ts function to convert the price variable to a time series.
Create a plot of partial autocorrelations of price.
Calculate a lag-1 price variable (note that the lag argument for the function is –1, not +1).
Create a scatterplot of price vs lag1price.
Use the ts.intersect function to create a dataframe containing price and lag1price.
Fit a simple linear regression model of price vs lag1price (a first-order autoregression model).
```{r google}
library(zoo)
google.ts <- read.zoo("./Data/google_stock.txt", format="%m/%d/%Y",
                      header=T)
plot(google.ts)
 
google <- read.table("./Data/google_stock.txt", header=T)
attach(google)
 
price.ts <- ts(price)
pacf(price.ts)
 
lag1price <- lag(price.ts, -1)
plot(price.ts ~ lag1price, xy.labels=F)
 
lagdata <- ts.intersect(price.ts, lag1price, dframe=T)
summary(lm(price.ts ~ lag1price, data=lagdata))
 
detach(google)
```

## Earthquakes (autoregression model)
Load the earthquakes data.
Create a time series plot of the data.
Use the ts function to convert the Quakes variable to a time series.
Create a plot of partial autocorrelations of Quakes.
Calculate lag-1, lag-2, and lag-3 Quakes variables.
Use the ts.intersect function to create a dataframe containing Quakes and the three lag variables.
Fit a multiple linear regression model of Quakes versus the three lag variables (a third-order autoregression model).
```{r earth}
earthquakes <- read.table("./Data/earthquakes.txt", header=T)
attach(earthquakes)
 
plot(Year, Quakes, type="b")
 
Quakes.ts <- ts(Quakes)
pacf(Quakes.ts)
 
lag1Quakes <- lag(Quakes.ts, -1)
lag2Quakes <- lag(Quakes.ts, -2)
lag3Quakes <- lag(Quakes.ts, -3)
 
lagdata <- ts.intersect(Quakes.ts, lag1Quakes, lag2Quakes, lag3Quakes, dframe=T)
summary(lm(Quakes.ts ~ lag1Quakes + lag2Quakes + lag3Quakes, data=lagdata))
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept)  6.44916    1.78646   3.610 0.000496 ***
# lag1Quakes   0.16424    0.10063   1.632 0.106049    
# lag2Quakes   0.07125    0.10128   0.703 0.483517    
# lag3Quakes   0.26928    0.09783   2.753 0.007110 ** 
# ---
# Residual standard error: 3.851 on 93 degrees of freedom
# Multiple R-squared:  0.1388,  Adjusted R-squared:  0.111 
# F-statistic: 4.997 on 3 and 93 DF,  p-value: 0.002942
 
detach(earthquakes)
```

## Blaisdell company (regression with autoregressive errors)
Load the blaisdell data.
Fit a simple linear regression model of comsales vs indsales.
Use the dwt function in the car package to conduct the Durbin-Watson test on the residuals.
Conduct the Ljung-Box test on the residuals.
Perform the Cochrane-Orcutt procedure to transform the variables.
Forecast comsales for period 21 when indsales are projected to be $175.3 million.
Perform the Hildreth-Lu procedure to transform the variables.
Perform the first differences procedure to transform the variables.
```{r blaisdell}
blaisdell <- read.table("./Data/blaisdell.txt", header=T)
attach(blaisdell)
 
model.1 <- lm(comsales ~ indsales)
summary(model.1)
#              Estimate Std. Error t value Pr(>|t|)    
# (Intercept) -1.454750   0.214146  -6.793 2.31e-06 ***
# indsales     0.176283   0.001445 122.017  < 2e-16 ***
 
# Durbin-Watson Test
library(car)
dwt(model.1)
# lag Autocorrelation D-W Statistic p-value
#   1       0.6260046     0.7347256       0
# Alternative hypothesis: rho != 0
 
# Ljung-Box Q Test
Box.test(residuals(model.1), lag = 1, type = "Ljung")
# Box-Ljung test
# data:  residuals(model.1)
# X-squared = 9.0752, df = 1, p-value = 0.002591
 
# Cochrane-Orcutt Procedure
res.ts <- ts(residuals(model.1))
lag1res <- lag(res.ts, -1)
lagdata1 <- ts.intersect(res.ts, lag1res)
acp <- coef(lm(res.ts ~ lag1res -1, data=lagdata1)) # 0.6311636
y.ts <- ts(comsales)
x.ts <- ts(indsales)
lag1y <- lag(y.ts, -1)
lag1x <- lag(x.ts, -1)
y.co <- y.ts-acp*lag1y
x.co <- x.ts-acp*lag1x
model.2 <- lm(y.co ~ x.co)
summary(model.2)
#              Estimate Std. Error t value Pr(>|t|)    
# (Intercept) -0.394111   0.167230  -2.357   0.0307 *  
# x.co         0.173758   0.002957  58.767   <2e-16 ***
 
dwt(model.2)
# lag Autocorrelation D-W Statistic p-value
#   1       0.1473569      1.650248   0.306
# Alternative hypothesis: rho != 0
 
b0 <- coef(model.2)[1]/(1-acp) # -1.068524
sqrt(vcov(model.2)[1,1])/(1-acp) # se = 0.4533986
b1 <- coef(model.2)[2] # 0.1737583
 
fit.20 <- b0+b1*indsales[20] # 28.76577
res.20 <- comsales[20]-fit.20 # 0.01422919
fit.21 <- b0+b1*175.3 # 29.3913
forecast.21 <- fit.21+acp*res.20 # 29.40028
 
# Hildreth-Lu Procedure
sse <- vector()
for(i in 1:90){
  y.hl = y.ts-(0.09+0.01*i)*lag1y
  x.hl = x.ts-(0.09+0.01*i)*lag1x
  sse[i] <- sum(residuals(lm(y.hl ~ x.hl))^2)
}
acp <- 0.09+0.01*which.min(sse) # 0.96
y.hl = y.ts-acp*lag1y
x.hl = x.ts-acp*lag1x
model.3 <- lm(y.hl ~ x.hl)
summary(model.3)
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept)  0.07117    0.05797   1.228    0.236    
# x.hl         0.16045    0.00684  23.458 2.18e-14 ***
 
dwt(model.3)
# lag Autocorrelation D-W Statistic p-value
#   1        0.116145      1.725439   0.548
# Alternative hypothesis: rho != 0
 
coef(model.3)[1]/(1-acp) # 1.77933
sqrt(vcov(model.3)[1,1])/(1-acp) # 1.449373
 
# First Differences Procedure
 
y.fd = y.ts-lag1y
x.fd = x.ts-lag1x
dwt(lm(y.fd ~ x.fd))
# lag Autocorrelation D-W Statistic p-value
#   1       0.1160548      1.748834    0.62
# Alternative hypothesis: rho != 0
 
model.4 <- lm(y.fd ~ x.fd -1)
summary(model.4)
#      Estimate Std. Error t value Pr(>|t|)    
# x.fd 0.168488   0.005096   33.06   <2e-16 ***
 
mean(comsales)-coef(model.4)[1]*mean(indsales) # -0.3040052
 
detach(blaisdell)
```

## Metal fabricator and vendor employees (regression with autoregressive errors)
Load the employee data.
Fit a simple linear regression model of metal vs vendor.
Create a scatterplot of the data with a regression line.
Create a scatterplot of the residuals vs time order.
Create a plot of partial autocorrelations of the residuals.
Use the dwt function in the car package to conduct the Durbin-Watson test on the residuals.
Perform the Cochrane-Orcutt procedure to transform the variables.
```{r metal}
employee <- read.table("./Data/employee.txt", header=T)
attach(employee)
 
model.1 <- lm(metal ~ vendor)
summary(model.1)
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept) 2.847911   3.299962   0.863    0.392    
# vendor      0.122442   0.009423  12.994   <2e-16 ***
 
plot(x=vendor, y=metal,
     panel.last = lines(sort(vendor), fitted(model.1)[order(vendor)]))
 
plot(x=time, y=residuals(model.1), type="b",
     panel.last = abline(h=0, lty=2))
 
pacf(residuals(model.1))
 
# Durbin-Watson Test
dwt(model.1)
# lag Autocorrelation D-W Statistic p-value
#   1        0.772038     0.3592396       0
# Alternative hypothesis: rho != 0
 
# Cochrane-Orcutt Procedure
res.ts <- ts(residuals(model.1))
lag1res <- lag(res.ts, -1)
lagdata1 <- ts.intersect(res.ts, lag1res)
acp <- coef(lm(res.ts ~ lag1res -1, data=lagdata1)) # 0.831385
y.ts <- ts(metal)
x.ts <- ts(vendor)
lag1y <- lag(y.ts, -1)
lag1x <- lag(x.ts, -1)
y.co <- y.ts-acp*lag1y
x.co <- x.ts-acp*lag1x
model.2 <- lm(y.co ~ x.co)
summary(model.2)
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept)  4.87560    0.78655   6.199 6.78e-08 ***
# x.co         0.04795    0.01300   3.688 0.000505 ***
 
coef(model.2)[1]/(1-acp) # 28.91557
sqrt(vcov(model.2)[1,1])/(1-acp) # se = 4.664789
 
detach(employee)
```
