---
title: "Lesson 04"
author: "Christopher A. Swenson (chris@cswenson.com)"
date: "11/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Alcohol consumption and muscle strength
Load the alcoholarm data.
Fit a simple linear regression model with y = strength and x = alcohol.
Display model results.
Display a scatterplot of the data with the simple linear regression line.
Display a residual plot with fitted values on the horizontal axis.
Display a residual plot with x = alcohol on the horizontal axis.
```{r alc}
alcoholarm <- read.table("./Data/alcoholarm.txt", header=T)
attach(alcoholarm)
 
model <- lm(strength ~ alcohol)
summary(model)
# Coefficients:
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept) 26.36954    1.20273  21.925  < 2e-16 ***
# alcohol     -0.29587    0.05105  -5.796 5.14e-07 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 3.874 on 48 degrees of freedom
# Multiple R-squared:  0.4117,  Adjusted R-squared:  0.3994 
# F-statistic: 33.59 on 1 and 48 DF,  p-value: 5.136e-07
 
plot(x=alcohol, y=strength,
     xlab="Lifetime consumption of alcohol", ylab="Deltoid muscle strength",
     panel.last = lines(sort(alcohol), fitted(model)[order(alcohol)]))
 
plot(x=fitted(model), y=residuals(model),
     xlab="Fitted values", ylab="Residuals",
     panel.last = abline(h=0, lty=2))
 
plot(x=alcohol, y=residuals(model),
     xlab="Lifetime consumption of alcohol", ylab="Residuals",
     panel.last = abline(h=0, lty=2))
 
detach(alcoholarm)
```

## Blood pressure
Load the bloodpress data.
Fit a simple linear regression model with y = BP and x = Age, display model results, and display a scatterplot of the data with the simple linear regression line.
Fit a simple linear regression model with y = BP and x = Weight, display model results, and display a scatterplot of the data with the simple linear regression line.
Fit a simple linear regression model with y = BP and x = Duration, display model results, and display a scatterplot of the data with the simple linear regression line.
Display a residual plot for the model using x = Age with Weight on the horizontal axis.
Fit a multiple linear regression model with y = BP, x1 = Age, and x2 = Weight.
Display a residual plot for the model using x1 = Age and x2 = Weight with Duration on the horizontal axis.
```{r bp}
bloodpress <- read.table("./Data/bloodpress.txt", header=T)
attach(bloodpress)
 
model.1 <- lm(BP ~ Age)
summary(model.1)
# Coefficients:
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept)  44.4545    18.7277   2.374  0.02894 * 
# Age           1.4310     0.3849   3.718  0.00157 **
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 4.195 on 18 degrees of freedom
# Multiple R-squared:  0.4344,  Adjusted R-squared:  0.403 
# F-statistic: 13.82 on 1 and 18 DF,  p-value: 0.001574
plot(x=Age, y=BP,
     xlab="Age (years)", ylab="Diastolic blood pressure (mm Hg)",
     panel.last = lines(sort(Age), fitted(model.1)[order(Age)]))
 
model.2 <- lm(BP ~ Weight)
summary(model.2)
# Coefficients:
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept)  2.20531    8.66333   0.255    0.802    
# Weight       1.20093    0.09297  12.917 1.53e-10 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 1.74 on 18 degrees of freedom
# Multiple R-squared:  0.9026,  Adjusted R-squared:  0.8972 
# F-statistic: 166.9 on 1 and 18 DF,  p-value: 1.528e-10
plot(x=Weight, y=BP,
     xlab="Weight (pounds)", ylab="Diastolic blood pressure (mm Hg)",
     panel.last = lines(sort(Weight), fitted(model.2)[order(Weight)]))
 
model.3 <- lm(BP ~ Dur)
summary(model.3)
# Coefficients:
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept) 109.2350     3.8563  28.327   <2e-16 ***
# Dur           0.7411     0.5703   1.299     0.21    
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 5.333 on 18 degrees of freedom
# Multiple R-squared:  0.08575,  Adjusted R-squared:  0.03496 
# F-statistic: 1.688 on 1 and 18 DF,  p-value: 0.2102
plot(x=Dur, y=BP,
     xlab="Duration of hypertension (years)",
     ylab="Diastolic blood pressure (mm Hg)",
     panel.last = lines(sort(Dur), fitted(model.3)[order(Dur)]))
 
plot(x=Weight, y=residuals(model.1),
     xlab="Weight (pounds)", ylab="Residuals from model with Age",
     panel.last = abline(h=0, lty=2))
 
model.12 <- lm(BP ~ Age + Weight)
 
plot(x=Dur, y=residuals(model.12),
     xlab="Duration of hypertension (years)",
     ylab="Residuals from model with Age and Weight",
     panel.last = abline(h=0, lty=2))
 
detach(bloodpress)
```

## Tread wear
Load the treadwear data.
Fit a simple linear regression model with y = groove and x = mileage.
Display model results.
Display a scatterplot of the data with the simple linear regression line.
Display a residual plot with fitted values on the horizontal axis.
```{r tread}
treadwear <- read.table("./Data/treadwear.txt", header=T)
attach(treadwear)
 
model <- lm(groove ~ mileage)
summary(model)
# Coefficients:
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept) 360.6367    11.6886   30.85 9.70e-09 ***
# mileage      -7.2806     0.6138  -11.86 6.87e-06 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 19.02 on 7 degrees of freedom
# Multiple R-squared:  0.9526,  Adjusted R-squared:  0.9458 
# F-statistic: 140.7 on 1 and 7 DF,  p-value: 6.871e-06
 
plot(x=mileage, y=groove,
     xlab="Mileage (1000s of miles)", ylab="Depth of groove (mils)",
     panel.last = lines(sort(mileage), fitted(model)[order(mileage)]))
 
plot(x=fitted(model), y=residuals(model),
     xlab="Fitted values", ylab="Residuals",
     panel.last = abline(h=0, lty=2))
 
detach(treadwear)
```

## Plutonium
Load the alphapluto data.
Fit a simple linear regression model with y = alpha and x = pluto.
Display model results.
Display a scatterplot of the data with the simple linear regression line.
Display a residual plot with fitted values on the horizontal axis.
```{r pluto}
alphapluto <- read.table("./Data/alphapluto.txt", header=T)
attach(alphapluto)
 
model <- lm(alpha ~ pluto)
summary(model)
# Coefficients:
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept) 0.0070331  0.0035988   1.954   0.0641 .  
# pluto       0.0055370  0.0003659  15.133 9.08e-13 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 0.01257 on 21 degrees of freedom
# Multiple R-squared:  0.916,  Adjusted R-squared:  0.912 
# F-statistic:   229 on 1 and 21 DF,  p-value: 9.077e-13
 
plot(x=pluto, y=alpha,
     xlab="Plutonium activity (pCi/g)", ylab="Alpha count rate (number per second)",
     panel.last = lines(sort(pluto), fitted(model)[order(pluto)]))
 
plot(x=fitted(model), y=residuals(model),
     xlab="Fitted values", ylab="Residuals",
     panel.last = abline(h=0, lty=2))
 
detach(alphapluto)
```

## Alcohol and tobacco
Load the alcoholtobacco data.
Fit a simple linear regression model with y = Alcohol and x = Tobacco.
Display model results.
Display a scatterplot of the data with the simple linear regression line.
Display a residual plot with fitted values on the horizontal axis.
Refit the model excluding Northern Ireland.
Display a scatterplot of the data excluding Northern Ireland with the simple linear regression line for the model excluding Northern Ireland.
Display a standardized residual plot for the model fit to all the data with fitted values on the horizontal axis.
Calculate the standardized residual for Northern Ireland.
```{r alc2}
alcoholtobacco <- read.table("./Data/alcoholtobacco.txt", header=T)
attach(alcoholtobacco)
 
model.1 <- lm(Alcohol ~ Tobacco)
summary(model.1)
# Coefficients:
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept)   4.3512     1.6067   2.708   0.0241 *
# Tobacco       0.3019     0.4388   0.688   0.5087  
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 0.8196 on 9 degrees of freedom
# Multiple R-squared:  0.04998,  Adjusted R-squared:  -0.05557 
# F-statistic: 0.4735 on 1 and 9 DF,  p-value: 0.5087
 
plot(x=Tobacco, y=Alcohol,
     xlab="Ave weekly tobacco expenditure (GBP)", 
     ylab="Ave weekly alcohol expenditure (GBP)",
     panel.last = lines(sort(Tobacco), fitted(model.1)[order(Tobacco)]))
 
plot(x=fitted(model.1), y=residuals(model.1),
     xlab="Fitted values", ylab="Residuals",
     panel.last = abline(h=0, lty=2))
 
model.2 <- lm(Alcohol ~ Tobacco, subset=Region!="NorthernIreland")
 
plot(x=Tobacco[Region!="NorthernIreland"], y=Alcohol[Region!="NorthernIreland"],
     xlab="Ave weekly tobacco expenditure (GBP)", 
     ylab="Ave weekly alcohol expenditure (GBP)",
     panel.last = lines(sort(Tobacco), fitted(model.2)[order(Tobacco)]))
 
plot(x=fitted(model.1), y=rstandard(model.1),
     xlab="Fitted values", ylab="Standardized residuals",
     panel.last = abline(h=0, lty=2))
 
rstandard(model.1)[Region=="NorthernIreland"] # -2.575075
 
detach(alcoholtobacco)
```

## Anscombe data
Load the anscombe data.
Fit a simple linear regression model with y = y3 and x = x3.
Display model results.
Display a scatterplot of the data with the simple linear regression line.
Display a residual plot with fitted values on the horizontal axis.
```{r ansc}
anscombe <- read.table("./Data/anscombe.txt", header=T)
attach(anscombe)
 
model <- lm(y3 ~ x3)
summary(model)
# Coefficients:
#             Estimate Std. Error t value Pr(>|t|)    
# (Intercept)   3.0025     1.1245   2.670  0.02562 * 
# x3            0.4997     0.1179   4.239  0.00218 **
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Residual standard error: 1.236 on 9 degrees of freedom
# Multiple R-squared:  0.6663,  Adjusted R-squared:  0.6292 
# F-statistic: 17.97 on 1 and 9 DF,  p-value: 0.002176
 
plot(x=x3, y=y3,
     panel.last = lines(sort(x3), fitted(model)[order(x3)]))
 
plot(x=fitted(model), y=residuals(model),
     xlab="Fitted values", ylab="Residuals",
     panel.last = abline(h=0, lty=2))
 
detach(anscombe)
```

## Skin cancer mortality
Load the skin cancer data.
Fit a simple linear regression model with y = Mort and x = Lat.
Display a scatterplot of the data with the simple linear regression line.
```{r skin}
skincancer <- read.table("./Data/skincancer.txt", header=T)
attach(skincancer)
 
model <- lm(Mort ~ Lat)
 
plot(x=Lat, y=Mort,
     xlab="Latitude (at center of state)", ylab="Mortality (deaths per 10 million)",
     main="Skin Cancer Mortality versus State Latitude",
     panel.last = lines(sort(Lat), fitted(model)[order(Lat)]))
 
detach(skincancer)
```

## Alligators
Load the alligator data.
Fit a simple linear regression model with y = weight and x = length.
Display a scatterplot of the data with the simple linear regression line.
```{r allig}
alligator <- read.table("./Data/alligator.txt", header=T)
attach(alligator)
 
model <- lm(weight ~ length)
 
plot(x=length, y=weight, ylim=c(-50, 650),
     panel.last = lines(sort(length), fitted(model)[order(length)]))
 
detach(alligator)
```

## Alloy corrosion
Load the corrosion data.
Fit a simple linear regression model with y = wgtloss and x = iron.
Display a scatterplot of the data with the simple linear regression line.
```{r alloy}
corrosion <- read.table("./Data/corrosion.txt", header=T)
attach(corrosion)
 
model <- lm(wgtloss ~ iron)
 
plot(x=iron, y=wgtloss,
     panel.last = lines(sort(iron), fitted(model)[order(iron)]))
 
detach(corrosion)
```

## Handcode and height
Load the handheight data.
Fit a simple linear regression model with y = HandSpan and x = Height.
Display a residual plot with fitted values on the horizontal axis.
```{r hand}
handheight <- read.table("./Data/handheight.txt", header=T)
attach(handheight)
 
model <- lm(HandSpan ~ Height)
 
plot(x=fitted(model), y=residuals(model),
     xlab="Fitted values", ylab="Residuals",
     panel.last = abline(h=0, lty=2))
 
detach(handheight)
```

## Chemical solution concentration
Load the solconc data.
Fit a simple linear regression model with y = y (concentration) and x = x (time).
Display a residual plot with fitted values on the horizontal axis.
```{r solconc}
solconc <- read.table("./Data/solutions_conc.txt", header=T)
attach(solconc)
 
model <- lm(y ~ x)
 
plot(x=fitted(model), y=residuals(model),
     xlab="Fitted values", ylab="Residuals",
     panel.last = abline(h=0, lty=2))
 
detach(solconc)
```

## Real estate sales
Load the realestate data.
Fit a simple linear regression model with y = SalePrice and x = Sqfeet.
Display a residual plot with fitted values on the horizontal axis.
```{r real}
realestate <- read.table("./Data/realestate_sales.txt", header=T)
attach(realestate)
 
model <- lm(SalePrice ~ SqFeet)
 
plot(x=fitted(model), y=residuals(model),
     xlab="Fitted values", ylab="Residuals",
     panel.last = abline(h=0, lty=2))
 
detach(realestate)
```

## Old Faithful geyser eruptions
Load the oldfaithful data.
Fit a simple linear regression model with y = waiting and x = eruption.
Display a histogram and normal probability plot of the residuals.
```{r old}
oldfaithful <- read.table("./Data/oldfaithful.txt", header=T)
attach(oldfaithful)

model <- lm(waiting ~ eruption)

hist(residuals(model), main="", breaks=12)
 
qqnorm(residuals(model), main="")
qqline(residuals(model))
 
detach(oldfaithful)
```

## Hospital infection risk
Load the infectionrisk data.
Select only hospitals in regions 1 or 2.
Fit a simple linear regression model with y = InfctRsk and x = Stay.
Display a normal probability plot of the residuals.
```{r inf}
infectionrisk <- read.table("./Data/infectionrisk.txt", header=T)
infectionrisk <- infectionrisk[infectionrisk$Region==1 | infectionrisk$Region==2, ]
attach(infectionrisk)
 
model <- lm(InfctRsk ~ Stay)
 
qqnorm(residuals(model), main="")
qqline(residuals(model))
 
detach(infectionrisk)
```

## Car stopping distances
Load the carstopping data.
Fit a simple linear regression model with y = Distance and x = Speed.
Display a scatterplot of the data with the simple linear regression line.
Display a residual plot with fitted values on the horizontal axis.
Create a new response variable equal to √Distance
Fit a simple linear regression model with y = √Distance and x = Speed.
Display a scatterplot of the data with the simple linear regression line.
Display a residual plot with fitted values on the horizontal axis.
Use the model to predict Distance for Speed = 10, 20, 30, and 40.
```{r car}
carstopping <- read.table("./Data/carstopping.txt", header=T)
attach(carstopping)
 
model <- lm(Distance ~ Speed)
plot(x=Speed, y=Distance,
     panel.last = lines(sort(Speed), fitted(model)[order(Speed)]))
plot(x=fitted(model), y=residuals(model),
     xlab="Fitted values", ylab="Residuals",
     panel.last = abline(h=0, lty=2))
 
sqrtdist <- sqrt(Distance)
 
model <- lm(sqrtdist ~ Speed)
plot(x=Speed, y=sqrtdist,
     panel.last = lines(sort(Speed), fitted(model)[order(Speed)]))
plot(x=fitted(model), y=residuals(model),
     xlab="Fitted values", ylab="Residuals",
     panel.last = abline(h=0, lty=2))
 
predict(model, interval="prediction",
        newdata=data.frame(Speed=c(10, 20, 30, 40)))^2
#         fit      lwr       upr
# 1  11.86090  3.93973  24.03997
# 2  35.63671 20.42935  55.04771
# 3  72.17067 49.44080  99.18664
# 4 121.46277 90.63292 156.79793
 
detach(carstopping)
```
