---
title: "Lesson 03"
author: "Christopher A. Swenson (chris@cswenson.com)"
date: "11/27/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Student heights and weights
Load the heightweight data.
Fit a simple linear regression model with y = wt and x = ht.
Display a scatterplot of the data with the simple linear regression line and a horizontal line at the mean weight.
Use the model to predict weight for height = 64.
```{r height}
heightweight <- read.table("./Data/student_height_weight.txt", header=T)
attach(heightweight)
 
model <- lm(wt ~ ht)
plot(x=ht, y=wt,
     panel.last = c(lines(sort(ht), fitted(model)[order(ht)]),
                    abline(h=mean(wt))))
mean(wt) # 158.8
predict(model, newdata=data.frame(ht=64)) # 126.2708
 
detach(heightweight)
```

## Skin cancer mortality
Load the skin cancer data.
Fit a simple linear regression model with y = Mort and x = Lat.
Display a scatterplot of the data with the simple linear regression line.
Use the model to calculate 95% confidence intervals for E(Mort) at Lat = 40 and 28.
Calculate mean(Lat).
Use the model to calculate 95% prediction intervals for Mort at Lat = 40.
Display a scatterplot of the data with the simple linear regression line, confidence interval bounds, and prediction interval bounds.
```{r skin}
skincancer <- read.table("./Data/skincancer.txt", header=T)
attach(skincancer)
 
model <- lm(Mort ~ Lat)
plot(x=Lat, y=Mort,
     xlab="Latitude (at center of state)", ylab="Mortality (deaths per 10 million)",
     main="Skin Cancer Mortality versus State Latitude",
     panel.last = lines(sort(Lat), fitted(model)[order(Lat)]))
 
predict(model, interval="confidence", se.fit=T,
        newdata=data.frame(Lat=c(40, 28)))
# $fit
#        fit      lwr      upr
# 1 150.0839 144.5617 155.6061
# 2 221.8156 206.8855 236.7456
# 
# $se.fit
# 1        2 
# 2.745000 7.421459 
 
mean(Lat) # 39.53265
 
predict(model, interval="prediction",
        newdata=data.frame(Lat=40))
#        fit     lwr      upr
# 1 150.0839 111.235 188.9329
 
plot(x=Lat, y=Mort,
     xlab="Latitude (at center of state)", ylab="Mortality (deaths per 10 million)",
     ylim=c(60, 260),
     panel.last = c(lines(sort(Lat), fitted(model)[order(Lat)]),
                    lines(sort(Lat), 
                          predict(model, 
                                  interval="confidence")[order(Lat), 2], col="green"),
                    lines(sort(Lat), 
                          predict(model, 
                                  interval="confidence")[order(Lat), 3], col="green"),
                    lines(sort(Lat), 
                          predict(model, 
                                  interval="prediction")[order(Lat), 2], col="purple"),
                    lines(sort(Lat), 
                          predict(model, 
                                  interval="prediction")[order(Lat), 3], col="purple")))
 
detach(skincancer)
```

## Hospital infection risk
Load the infectionrisk data.
Select only hospitals in regions 1 or 2.
Display a scatterplot of Stay versus InfctRsk.
Select only hospitals with Stay < 16 (i.e., remove the two hospitals with extreme values of Stay).
Fit a simple linear regression model with y = InfctRsk and x = Stay.
Use the model to calculate 95% confidence intervals for E(InfctRsk) at Stay = 10.
Use the model to calculate 95% prediction intervals for InfctRsk at Stay = 10.
Display a scatterplot of the data with the simple linear regression line, confidence interval bounds, and prediction interval bounds.
```{r infect}
infectionrisk <- read.table("./Data/infectionrisk.txt", header=T)
infectionrisk <- infectionrisk[infectionrisk$Region==1 | infectionrisk$Region==2, ]
attach(infectionrisk)
plot(x=Stay, y=InfctRsk)
detach(infectionrisk)
infectionrisk <- infectionrisk[infectionrisk$Stay<16, ]
attach(infectionrisk)
plot(x=Stay, y=InfctRsk)
 
model <- lm(InfctRsk ~ Stay)
 
predict(model, interval="confidence",
        newdata=data.frame(Stay=10))
#        fit      lwr      upr
# 1 4.528846 4.259205 4.798486
 
predict(model, interval="prediction",
        newdata=data.frame(Stay=10))
#        fit     lwr      upr
# 1 4.528846 2.45891 6.598781
 
plot(x=Stay, y=InfctRsk,
     ylim=c(0, 9),
     panel.last = c(lines(sort(Stay), fitted(model)[order(Stay)]),
                    lines(sort(Stay), 
                          predict(model, 
                                  interval="confidence")[order(Stay), 2], col="green"),
                    lines(sort(Stay), 
                          predict(model, 
                                  interval="confidence")[order(Stay), 3], col="green"),
                    lines(sort(Stay), 
                          predict(model, 
                                  interval="prediction")[order(Stay), 2], col="purple"),
                    lines(sort(Stay), 
                          predict(model, 
                                  interval="prediction")[order(Stay), 3], col="purple")))
 
detach(infectionrisk)
```
